---
title: 'GFM: installation and simulated example'
author: "Wei Liu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GFM Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



## Install the GFM
This vignette provides an introduction to the R package `GFM`, where the function `gfm` implements the model `GFM`, Generalized Factor Model for ultra-high dimensional variables with mixed types.
The package can be installed with the command:

`library(remotes)`

`remotes::install_github("feiyoung/GFM")`

or

`install.packages("GFM")`

The package can be loaded with the command:
```{r  eval=TRUE}
library("GFM")
```

## Fit GFM model using simulated data
### Generate data with  homogeneous  normal variables
First, we generate the data with  homogeneous  normal variables.
```{r  eval=TRUE}
## Homogeneous  normal variables
  dat <- gendata(q = 2, n=100, p=100, rho=3)
```
Then, we set the algorithm parameters and fit model
```{r}
# full-one vector means there is only one variable type within all variables.
  group <- rep(1,ncol(dat$X)) 
# set variables' type, 'gaussian' means there is  continous variable type.
  type <- 'gaussian' 
```
Third, we fit the GFM model with user-specified number of factors.
```{r}
# specify q=2
  gfm1 <- gfm(dat$X, group, type, q=2, output = FALSE)
  
  # measure the performance of GFM estimators in terms of canonical correlations
  measurefun(gfm1$hH, dat$H0, type='ccor')
  measurefun(gfm1$hB, dat$B0, type='ccor')
```
The number of factors can also be determined by data-driven manners.
```{r}
# select q automatically
  gfm2 <- gfm(dat$X, group, type, q=NULL, q_set = 1:6, output = FALSE)
# measure the performance of GFM estimators in terms of canonical correlations
  measurefun(gfm2$hH, dat$H0, type='ccor')
  measurefun(gfm2$hB, dat$B0, type='ccor')
```


### Generate data with  heterogeous normal variables
First, we generate the data with  heterogeous  normal variables and set the parameters of algorithm.
```{r  eval=TRUE}
  dat <- gendata(seed=1, n=100, p=100, type='heternorm', q=2, rho=1)
  group <- rep(1,ncol(dat$X))
  type <- 'gaussian'
```

Third, we fit the GFM model with user-specified number of factors and compare the results with that of linear factor models.
```{r}
# specify q=2
  gfm1 <- gfm(dat$X, group, type, q=2, output = FALSE)
  
  # measure the performance of GFM estimators in terms of canonical correlations
  corH_gfm <- measurefun(gfm1$hH, dat$H0, type='ccor')
  corB_gfm <- measurefun(gfm1$hB, dat$B0, type='ccor')
  
  lfm1 <- Factorm(dat$X, q=2)
  corH_lfm <- measurefun(lfm1$hH, dat$H0, type='ccor')
  corB_lfm <- measurefun(lfm1$hB, dat$B0, type='ccor')
  
  library(ggplot2)
  df1 <- data.frame(CCor= c(corH_gfm, corH_lfm, corB_gfm, corB_lfm),
                    Method =factor(rep(c('GFM', "LFM"), times=2)),
                    Quantity= factor(c(rep('factors',2), rep("loadings", 2))))
  ggplot(data=df1, aes(x=Quantity, y=CCor, fill=Method)) + geom_bar(position = "dodge", stat="identity",width = 0.5)
```

The number of factors can also be determined by data-driven manners.
```{r}
# select q automatically
  gfm2 <- gfm(dat$X, group, type, q=NULL, q_set = 1:4, output = FALSE)
# measure the performance of GFM estimators in terms of canonical correlations
  corH_gfm <- measurefun(gfm2$hH, dat$H0, type='ccor')
  corB_gfm <- measurefun(gfm2$hB, dat$B0, type='ccor')
  
  library(ggplot2)
  df1 <- data.frame(CCor= c(corH_gfm, corH_lfm, corB_gfm, corB_lfm),
                    Method =factor(rep(c('GFM', "LFM"), times=2)),
                    Quantity= factor(c(rep('factors',2), rep("loadings", 2))))
  ggplot(data=df1, aes(x=Quantity, y=CCor, fill=Method)) + geom_bar(position = "dodge", stat="identity",width = 0.5)
```

### Generate data with  Count(Poisson) variables
First, we generate the data with  Count(Poisson) variables and set the parameters of algorithm.
```{r  eval=TRUE}
  q <- 3; p <- 200
  dat <- gendata(seed=1, n=200, p=p, type='pois', q=q, rho=4)
  group <- rep(1,ncol(dat$X))
  type <- 'poisson'
```

Second, we we fit the GFM models in the parallel manner. 
```{r}
  system.time(
    gfm2 <- gfm(dat$X, group, type,parallel = TRUE, q=NULL, q_set = 1:4, output = FALSE, fast_version = TRUE))
```

Third, we compare the results with that of linear factor models.
```{r}

  # measure the performance of GFM estimators in terms of canonical correlations
  corH_gfm <- measurefun(gfm2$hH, dat$H0, type='ccor')
  corB_gfm <- measurefun(gfm2$hB, dat$B0, type='ccor')
  
  lfm1 <- Factorm(dat$X, q=3)
  corH_lfm <- measurefun(lfm1$hH, dat$H0, type='ccor')
  corB_lfm <- measurefun(lfm1$hB, dat$B0, type='ccor')
  
  library(ggplot2)
  df1 <- data.frame(CCor= c(corH_gfm, corH_lfm, corB_gfm, corB_lfm),
                    Method =factor(rep(c('GFM', "LFM"), times=2)),
                    Quantity= factor(c(rep('factors',2), rep("loadings", 2))))
  ggplot(data=df1, aes(x=Quantity, y=CCor, fill=Method)) + geom_bar(position = "dodge", stat="identity",width = 0.5)
```


### Generate data with  the mixed-types of Poisson and Binomial variables
First, we generate the data with  Count(Poisson) variables and set the parameters of algorithm. Then fit the GFM model with user-specified number of factors.
```{r  eval=TRUE}
  dat <- gendata(seed=1, n=200, p=200, type='pois_bino', q=2, rho=2)
  group <- c(rep(1,ncol(dat$X)/2), rep(2,ncol(dat$X)/2))
  type <- c('poisson','binomial')
  table(dat$X[,1])
  table(dat$X[, 200])
  # user-specified q=2
  gfm2 <- gfm(dat$X, group, type, dropout = 2, q=2, output = FALSE, maxIter=5)
  measurefun(gfm2$hH, dat$H0, type='ccor')
  measurefun(gfm2$hB, dat$B0, type='ccor')
```


Third, we compare the results with that of linear factor models.
```{r}
  #  select q automatically
  gfm2 <- gfm(dat$X, group, type, dropout = 2, q=NULL, q_set = 1:4, output = FALSE)
  # measure the performance of GFM estimators in terms of canonical correlations
  corH_gfm <- measurefun(gfm2$hH, dat$H0, type='ccor')
  corB_gfm <- measurefun(gfm2$hB, dat$B0, type='ccor')
  
```
Compare with linear factor models
```{r}
  lfm1 <- Factorm(dat$X, q=3)
  corH_lfm <- measurefun(lfm1$hH, dat$H0, type='ccor')
  corB_lfm <- measurefun(lfm1$hB, dat$B0, type='ccor')
  
  library(ggplot2)
  df1 <- data.frame(CCor= c(corH_gfm, corH_lfm, corB_gfm, corB_lfm),
                    Method =factor(rep(c('GFM', "LFM"), times=2)),
                    Quantity= factor(c(rep('factors',2), rep("loadings", 2))))
  ggplot(data=df1, aes(x=Quantity, y=CCor, fill=Method)) + geom_bar(position = "dodge", stat="identity",width = 0.5)
```
## Session information
```{r}
sessionInfo()
```
